\\Large Rising temperatures alter how and where mammals move

### 2025-09-27 writing/manuscript.Rmd
started addressing Mike's comments

# Alternative titles {-}

* Rising temperatures alter the movement phenology of terrestrial mammals
* Terrestrial mammals alter their movement behaviour to rising temperatures

@@

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but many did not disentangle them from seasonal behaviour cycles. Consequently, it is still unclear how mammals will adapt their movement behaviour and phenology in the 21^st^ century. We address this gap by using hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023. We show that temperature significantly affects of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century. However, species did not respond uniformly to rising temperatures, so community-level responses will likely be complex, as climate change will cause species to alter their seasonal movement behavior and increase their chances of hyperythermia. Median habitat selection strength decreased for most species, but some species showed increased selection strength for higher altitudes. Consequently, our results suggest that all six species may shift their range in the coming decades as they begin to select against their current ranges. As climate change exposes mammals to novel environmental conditions, proactively predicting changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and protecting high-quality habitat. We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can reduce movement rates and cause physiological damage [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016]. By altering mammals' movement speed and habitat use [jointly, their movement behaviour: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behaviour [@alston_temperature_2020; @giroux_activity_2023], such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. Behavioural changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].

@@

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024]. Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, especially with increasing body size [@dyer_travel_2023; @verzuh_behavioural_2023]. Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature generalists') respond more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond to within a lifetime or a few generations [@levins_evolution_1974; @gienapp_climate_2008; @fuller_towards_2016; @xu_plasticity_2021].

@@

In the last century, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023]. As heat stress intensifies over the next 100 years, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020], which will be exacerbated by more frequent and intense extreme events [@bunnell_global_2011] and anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Preparing for and responding to future changes require a detailed understanding of how temperature will affect how and where different mammalian species move [@cunningham_opportunity_2021]. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021], and movement behaviour [@mccain_assessing_2019; @melin_effect_2023; @cunningham_opportunity_2021]. However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of changes in movement behaviour with climate change in the current century [@ipcc_climate_2023].

<!-- zoom in to western canada -->

Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023]. As we approach the deadline for the Kunming-Montreal Global Biodiversity Framework (i.e., the "30 by 30" initiative), a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].

@@

Large-bodied mammals are most likely to show changes in movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023]. Conveniently, their large body size also allows for larger tracking devices that have longer battery life and allow longer tracking periods and with finer temporal resolution [@kenward_manual_2001]. In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou: *Rangifer tarandus*, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves: *Canis lupus*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of moving, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with monthly climate change projections [@mahony_global_2022] to forecast each species' movement behaviour during the 21^st^ century under four different climate-change scenarios, namely Shared Socioeconomic Pathways (SSPs): 1-2.6 ("Sustainability -- Taking the Green Road"; low challenges to mitigation and adaptation), 2-4.5 ("Middle of the Road: medium challenges to mitigation and adaptation), 3-7.0 ("Regional Rivalry -- A Rocky Road"; high challenges to mitigation and adaptation), and 5-8.5 ("Fossil-fueled Development -- Taking the Highway"; high challenges to mitigation, low challenges to adaptation). @riahi_shared_2017 provide a detailed overview of SSPs. We discuss the consequences of these changes in movement behaviour at the individual, population, and community levels. We also provide directions for future research and solutions for overcoming some common data limitations.

@@

This study leverages four main datasets: (1) a multi-species collection of GPS telemetry data (Fig. \@ref(fig:bc-map)); (2) historical hourly air temperature from the ERA5 reanalyzed dataset [European Center for Medium-range Weather Forecasting$;$ @hersbach_era5_2023]; (3) temporally static rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs$;$ @riahi_shared_2017]. While we acknowledge there are other important variables (e.g., forest type, structure, and age; terrain slope and aspect), we limited our models to only use three simple variables for the sake of parsimony and comparability across species. The simplicity of the models also allowed for simple forecasts under the assumption of temporally invariant landscapes. See the discussion section for additional detail on the advantages and limitations of using such simple models.

@@

\noindent Boreal caribou and wolf telemetries were acquired from a data repository of the British Columbia Oil and Gas Research and Innovation Society (BC ORGIS) that was collected as part of the boreal caribou radio-collaring program of the BC OGRIS Research and Effectiveness Monitoring Board [REMB$;$ @bc_ogris_boreal_2018]. Southern mountain caribou data ***XXX***. Cougar data include telemetries from @darlington_southern_2025 and @serrouya_ref. Elk data from the work of @ciuti_human_2012 were downloaded from Movebank [@kays_movebank_2022]. ***MISSING INFO ON GRIZZLY BEAR AND GOAT TELEMETRY DATA.***

@@

\noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 database [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)). Proximal air temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [v. 1.7-46, @hijmans_terra_2023] for `R` [@r_core_team_r_2024].

@@

\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`, in meters), and distance from water (`dist_water_m`, in meters) by fitting a Hierarchical Resource Selection Function (HRSF) for each species. We fit each HRSF using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008]. After removing non-resident individuals (Table B1), we accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019; @alston_mitigating_2022] to produce estimates of second-order habitat selection [@johnson_comparison_1980]. Quadrature points were determined using the raster cells in each animal's 99.9% AKDE percentile and given a weight of 1. The number of quadrature locations greatly outnumbered the number of observed locations (Fig. B12), especially after accounting for the AKDE weights (Fig. B13).

@@

Changes in mammals' seasonal movement behaviour are also likely to affect the timing and frequency of human-wildlife interactions. Since heat affects mammals' sleep quality and likelihood to enter torpor [@mortlock_sleep_2024; @fjelldal_nightly_2021; @wells_life_2022], future decades may see increased levels of incidents with bears that struggle to enter or remain in hibernation [@johnson_human_2018]. The risk is further increased when bears' are unable to meet their increased energetic requirements through foraging or hunting. Projected increases in park attendance due to rising temperatures and human population density in Western Canada are expected to further increase frequency of human-wildlife interactions [@weststrate_how_2024].

@@

Our results demonstrate that mammals respond to changes in temperature by altering how and where they move in complex and nonlinear ways (Figs. \@ref(fig:dist) and \@ref(fig:hrsfs)). Hierarchical Generalized Additive Models were crucial in producing such results. The hierarchical approach us to produce statistically sound population-level estimates by accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mortlock_sleep_2024; @jeltsch_need_2025; @silva_too_2025], while the flexibility of GAMs estimated mammals' changes in movement behaviour without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018]. Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously by inexperienced analysts. We suggest choosing the basis size (`k`) for each smooth term carefully. While penalized splines and REML can help avoid unruly smooth terms that would result in questionable estimates the penalty terms can struggle to restrict model complexity when datasets are particularly large [@wood_generalized_2017; @wood_generalized_2017-1].

@@

We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection in hopes that other researchers can leverage this framework and models to expand our understanding of how temperature affects the movement behavior of other species, including smaller mammals, more elusive species, and non-mammal animals.

>>>>

@@

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but few disentangled them from seasonal behaviour cycles. Consequently, it is still uncertain how mammals will adapt their movement behaviour throughout the 21^st^ century. We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023. We show that temperature significantly affects of how and where these mammals move, and we predict that climate change will have significant impacts on mammals' movement behaviour throughout the 21^st^ century. Notably, we found that species did not respond uniformly to temperature. This suggests that community-level responses will be complex, as future climate change drives some species to move more and others less. Median habitat selection strength was predicted to decrease for most species, but some species are expected to show increased selection strength for higher altitudes. Consequently, our results suggest that individuals from all six species may shift their range in the coming decades as they begin to select against portions of their current ranges. The strong but varied responses to temperature suggest that future climate change will impact not only individuals' movement behaviour but also entire communities, as changes in movement behaviour will affect encounter rates, including predator prey-dynamics and human-wildlife interactions. As climate change exposes mammals to novel environmental conditions, proactively anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. How and where animals move thus determines encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can reduce movement rates [@dyer_travel_2023] and even cause physiological damage and death [@hetem_responses_2014; @ratnayake_forecasting_2019; @schmidt_interplay_2020]. Animals may search for a short-term buffer from extreme heat or cold by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016]. Changes in climate throughout the 21^st^ century and the simultaneous increase in frequency and intensity of extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011] are expected to impact not only animals' movement behavior but also their physiology [@powers_hovering_2017; @schwerdt_ecological_2024], life history [@newediuk_climate_2024], and phenology [@carbeck_adaptation_2022; @johnson_human_2018].

@@

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024]. Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, especially with increasing body size. Large-bodied mammals are most likely to show changes in movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023].<!-- Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature generalists') respond more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond to within a lifetime or a few generations [@levins_evolution_1974; @gienapp_climate_2008; @fuller_towards_2016; @xu_plasticity_2021].-->

@@

In the last century, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023]. As heat stress intensifies over the next 100 years, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020], which will be exacerbated by more frequent and intense extreme events [@bunnell_global_2011] and anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Preparing for and responding to future changes will require a detailed understanding of how temperature will affect how and where different mammalian species move [@cunningham_opportunity_2021]. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021], and movement behaviour [@mccain_assessing_2019; @melin_effect_2023; @cunningham_opportunity_2021]. However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of changes in movement behaviour with climate change in the current century [@ipcc_climate_2023].

@@

***CLEAN UP PARAGRAPH BELOW***

In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou: *Rangifer tarandus*, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves: *Canis lupus*; Table \@ref(tab:data-table)). Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme heat events [@zhang_increased_2023]. Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024]. We discuss the consequences of our findings at the individual, population, and community levels, and we also provide directions for future research and solutions for overcoming some common data limitations.

@@

In this study, we estimate how temperature affected mammals' probability of moving, speed when moving, and habitat selection by leveraging three datasets: (1) a multi-species collection of 25 years of GPS telemetry data throughout a large spatial range of western Canada (Fig. \@ref(fig:bc-map)), (2) historical, hourly air temperature data from the ERA5 reanalyzed dataset [@hersbach_era5_2023], and (3) temporally static rasters of percent forest cover, elevation, and distance from water. We then forecast each species' movement behaviour thoughout the 21^st^ century using monthly climate change projections under four Shared Socioeconomic Pathways [SSPs$;$ @riahi_shared_2017; @mahony_global_2022]: SSP 1-2.6 ("Sustainability -- Taking the Green Road"; low challenges to mitigation and adaptation), SSP 2-4.5 ("Middle of the Road: medium challenges to mitigation and adaptation), SSP 3-7.0 ("Regional Rivalry -- A Rocky Road"; high challenges to mitigation and adaptation), and SSP 5-8.5 ("Fossil-fueled Development -- Taking the Highway"; high challenges to mitigation, low challenges to adaptation).

We acknowledge that our models of habitat selection ignore many important variables of habitat selection (e.g., forest type, structure, and age; terrain slope and aspect; prey availability; population density). We limited our models to only use three simple and temporally static rasters to produce simple and easily interpretable estimates and forecasts under the assumption of temporally invariant landscapes. See the discussion section for additional detail on the advantages and limitations of using such simple models.

@@

\noindent Boreal caribou and wolf telemetries were acquired from a data repository of the British Columbia Oil and Gas Research and Innovation Society (BC ORGIS) that was collected as part of the boreal caribou radio-collaring program of the BC OGRIS Research and Effectiveness Monitoring Board [REMB$;$ @bc_ogris_boreal_2018]. Southern mountain caribou data were obtained from the work of @ford_effects_2023. Cougar data include telemetries from @darlington_southern_2025 and @serrouya_ref. Elk data from the work of @ciuti_human_2012 were downloaded from Movebank [@kays_movebank_2022]. ***MISSING INFO ON GRIZZLY BEAR AND GOAT TELEMETRY DATA.***

@@

\noindent Rasters of hourly air temperature data were downloaded from the ERA5 database [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)). Proximal air temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [v. 1.7-46, @hijmans_terra_2023] for `R` [@r_core_team_r_2024].

@@

\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`, in meters), and distance from water (`dist_water_m`, in meters) by fitting a Hierarchical Resource Selection Function (HRSF) for each species [@mccabe_resource_2021]. We fit each HRSF using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008]. After removing non-resident individuals (Table B1), we accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019; @alston_mitigating_2022] to produce estimates of second-order habitat selection [@johnson_comparison_1980]. Quadrature points were determined using the raster cells in each animal's 99.9% AKDE percentile and given a weight of 1. The number of quadrature locations greatly outnumbered the number of observed locations (Fig. B12), especially after accounting for the AKDE weights (Fig. B13).

@@

Changes in mammals' seasonal movement behaviour are also likely to affect the timing and frequency of human-wildlife interactions [@abrahms_climate_2023; @weststrate_how_2024]. Since heat affects mammals' sleep quality and likelihood to enter torpor [@mortlock_sleep_2024; @fjelldal_nightly_2021; @wells_life_2022], future decades may see increased levels of incidents with bears that struggle to enter or remain in hibernation [@johnson_human_2018]. The risk is further increased when bears' are unable to meet their increased energetic requirements through foraging or hunting. Projected increases in park attendance due to rising temperatures and human population density in Western Canada are expected to further increase frequency of human-wildlife interactions [@weststrate_how_2024].

@@

Our results demonstrate that mammals respond to changes in temperature by altering how and where they move in complex and nonlinear ways (Figs. \@ref(fig:dist) and \@ref(fig:hrsfs)). Hierarchical Generalized Additive Models were crucial in producing such results. The hierarchical approach us to produce statistically sound population-level estimates by accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @mortlock_sleep_2024; @jeltsch_need_2025; @silva_too_2025], while the flexibility of GAMs estimated mammals' changes in movement behaviour without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018]. Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously by inexperienced analysts. We suggest choosing the basis size (`k`) for each smooth term carefully. While penalized splines and REML can help avoid unruly smooth terms that would result in questionable estimates the penalty terms can struggle to restrict model complexity when datasets are particularly large [@wood_generalized_2017; @wood_generalized_2017-1].

@@

We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection in hopes that other researchers can leverage this framework and models to expand our understanding of how temperature affects the movement behaviour of other species, including smaller mammals, more elusive species, and non-mammal animals.
